---
title: "Compulsory Assignment 2"
subtitle: "Group 97 - Predictive Analytics (KAN-CDSCO1005U)"
author: "Alejandro Gato (161989), Alexander Myrup (160363), Pontus Neumann (185912), Linus Yu (160714)"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

\newpage

# Part 1

## 1.1 Time Series Plot, ACF, and Decomposition

```{r load-packages}
library(forecast)
library(tseries)
library(ggplot2)
library(MetBrewer)
library(gridExtra)
```

```{r load-data}
raw <- read.csv("data/fcpi.csv", sep = ";")

fcpi <- ts(raw$cpi, start = c(1960, 1), frequency = 12)

bene <- met.brewer("Benedictus")
```

```{r ts-plot, fig.height=4}
autoplot(fcpi) +
  geom_line(color = bene[1], linewidth = 0.7) +
  labs(
    title = "French Consumer Price Index (1960 to 2019)",
    x     = "Year",
    y     = "CPI"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    panel.grid.minor = element_blank()
  )
```

The French CPI series displays a strong upward trend across the full sample period with no signs of mean reversion. The level rises from approximately 10 in 1960 to over 100 by 2019. Growth is steeper in the 1970s and early 1980s, likely driven by the oil crises and accompanying inflation, and flattens noticeably from the mid 1980s onward. A mild seasonal pattern is visible, particularly in the later decades. The variance of the series increases with its level, suggesting that a variance stabilising transformation may be warranted.

```{r acf-plot, fig.height=4}
acf_data <- acf(as.numeric(fcpi), lag.max = 48, plot = FALSE)
acf_df   <- data.frame(
  lag = as.numeric(acf_data$lag[-1]),
  acf = as.numeric(acf_data$acf[-1])
)
ci <- qnorm(0.975) / sqrt(length(fcpi))

ggplot(acf_df, aes(x = lag, y = acf)) +
  geom_hline(yintercept = 0, color = "grey40") +
  geom_hline(yintercept = c(-ci, ci), linetype = "dashed", color = bene[3]) +
  geom_segment(aes(xend = lag, yend = 0), color = bene[1], linewidth = 0.6) +
  geom_point(color = bene[1], size = 1.5) +
  labs(
    title = "ACF of French CPI",
    x     = "Lag",
    y     = "Autocorrelation"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    panel.grid.minor = element_blank()
  )
```

The ACF decays extremely slowly, remaining significant well beyond 48 lags. This is characteristic of a non-stationary series with a strong trend component. The persistence confirms that the raw CPI level must be differenced or detrended before any ARIMA modelling.

```{r decomposition, fig.height=6}
decomp <- decompose(fcpi, type = "multiplicative")
autoplot(decomp) +
  labs(title = "Classical Multiplicative Decomposition of French CPI") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    panel.grid.minor = element_blank()
  )
```

A classical multiplicative decomposition is chosen because the seasonal amplitude appears to grow with the level of the series. The decomposition reveals: (i) a smooth upward trend that captures the long-run inflationary movement, (ii) a seasonal component with a regular annual pattern, and (iii) an irregular (remainder) component that is largest during the high inflation period of the 1970s.

**Limitations of classical decomposition.** The method uses symmetric moving averages, so the trend is unavailable for the first and last six months. It assumes that the seasonal pattern is perfectly constant over time, which is unlikely across a 60 year span where the structure of the economy changed considerably. It also offers no built-in mechanism for handling outliers or structural breaks, meaning that large shocks (e.g. the oil crises) feed directly into the remainder without adjustment.

## 1.2 Transformation

```{r boxcox-lambda}
lambda <- BoxCox.lambda(fcpi)
cat("Optimal Box-Cox lambda:", round(lambda, 4), "\n")
```

```{r transform-compare, fig.height=6}
fcpi_log <- log(fcpi)
fcpi_bc  <- BoxCox(fcpi, lambda)

p1 <- autoplot(fcpi_log) +
  geom_line(color = bene[2], linewidth = 0.6) +
  labs(title = "Log Transformed CPI", x = "Year", y = "log(CPI)") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    panel.grid.minor = element_blank()
  )

p2 <- autoplot(fcpi_bc) +
  geom_line(color = bene[4], linewidth = 0.6) +
  labs(
    title = paste0("Box-Cox Transformed CPI (lambda = ", round(lambda, 2), ")"),
    x     = "Year",
    y     = "Transformed CPI"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    panel.grid.minor = element_blank()
  )

grid.arrange(p1, p2, ncol = 1)
```

```{r decomp-log, fig.height=6}
decomp_log <- decompose(fcpi_log, type = "additive")
autoplot(decomp_log) +
  labs(title = "Classical Additive Decomposition of log(CPI)") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    panel.grid.minor = element_blank()
  )
```

After the log transformation, the variance of the series is substantially more stable over time, and the seasonal fluctuations become approximately constant in magnitude. This means an additive decomposition is now appropriate. The trend component remains strongly upward but is smoother in its curvature, and the remainder is more homoscedastic than in the original series.

The Box-Cox transformation yields a lambda close to zero, confirming that the log transformation is a near optimal variance stabilising choice. Because lambda $\approx 0$ and the natural logarithm is simpler to interpret (percentage changes after differencing), the **log transformed series is used for the remainder of the assignment**.

## 1.3 Unit Root Tests (KPSS and ADF)

The log transformed CPI displays both a clear deterministic trend and a level shift, so the appropriate test specifications include a trend term.

- **KPSS**: use `type = "Trend"` to test stationarity around a deterministic trend.
- **ADF**: use `k = 12` (monthly lag length) to account for serial correlation, with the alternative being trend stationarity.

```{r unit-root-tests}
kpss_result <- kpss.test(fcpi_log, null = "Trend")
adf_result  <- adf.test(fcpi_log, k = 12)

ur_tbl <- data.frame(
  Test      = c("KPSS (Trend)", "ADF"),
  Statistic = c(round(kpss_result$statistic, 4),
                round(adf_result$statistic, 4)),
  p.value   = c(format.pval(kpss_result$p.value, digits = 3, eps = 0.01),
                format.pval(adf_result$p.value, digits = 3, eps = 0.01))
)

knitr::kable(
  ur_tbl,
  col.names = c("Test", "Test Statistic", "p-value"),
  caption   = "Unit root tests on log(CPI).",
  align     = "lcc"
)
```

**KPSS** tests the null of (trend )stationarity. The p-value is below 0.01, so the null is rejected: the series is not stationary even after accounting for a deterministic trend. **ADF** tests the null of a unit root. The p-value is large, so the null cannot be rejected. Both tests consistently indicate that log(CPI) contains a unit root and requires differencing.

## 1.4 Making the Series Stationary and ARIMA Identification

Since the series contains both a trend (unit root) and seasonality, first order regular differencing and first order seasonal differencing are applied.

```{r differencing, fig.height=4}
fcpi_d1  <- diff(fcpi_log, differences = 1)
fcpi_d1s <- diff(fcpi_d1, lag = 12, differences = 1)

autoplot(fcpi_d1s) +
  geom_line(color = bene[2], linewidth = 0.5) +
  labs(
    title = "Differenced log(CPI): First Regular + First Seasonal Difference",
    x     = "Year",
    y     = expression(Delta[1] * Delta[12] ~ log(CPI))
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    panel.grid.minor = element_blank()
  )
```

```{r kpss-differenced}
kpss_d <- kpss.test(fcpi_d1s, null = "Level")
cat("KPSS test on differenced series: statistic =", round(kpss_d$statistic, 4),
    ", p-value =", format.pval(kpss_d$p.value, digits = 3, eps = 0.01), "\n")
```

After both regular and seasonal differencing, the series fluctuates around zero with no visible trend or changing variance. The KPSS test on the differenced series does not reject the null of level stationarity, confirming that the series is now stationary.

```{r acf-pacf, fig.height=6}
par(mfrow = c(2, 1), mar = c(4, 4, 3, 1))
Acf(fcpi_d1s, lag.max = 48, main = "ACF of Differenced log(CPI)")
Pacf(fcpi_d1s, lag.max = 48, main = "PACF of Differenced log(CPI)")
par(mfrow = c(1, 1))
```

**Reading the ACF and PACF:**

- The ACF shows a significant spike at lag 1 that cuts off afterwards, suggesting a non-seasonal MA(1) component.
- At the seasonal lags (12, 24, ...), the ACF shows a significant spike at lag 12 that decays, suggesting a seasonal MA(1) component.
- The PACF shows a significant spike at lag 1 with subsequent decay, consistent with the MA interpretation rather than an AR process.

Based on this, the proposed model is:

$$\text{ARIMA}(0,1,1)(0,1,1)_{12}$$

This is the classic "airline model" (Box and Jenkins), which is well suited to economic series displaying a trend and multiplicative seasonality. In differenced form, the model can be written as:

$$(1 - B)(1 - B^{12}) \log(Y_t) = (1 + \theta_1 B)(1 + \Theta_1 B^{12}) \varepsilon_t$$

where $B$ is the backshift operator, $\theta_1$ is the non-seasonal MA coefficient, and $\Theta_1$ is the seasonal MA coefficient.

\newpage

# Part 2

In this part, the log-transformed series is used throughout, as decided in 1.2.

## 2.1 Train/Test Split and Auto ARIMA

The sample is split so that the last 24 months (Jan 2018 to Dec 2019) serve as the test set. This provides two full years of out-of-sample observations while retaining 58 years of training data.

```{r train-test-split}
train <- window(fcpi_log, end = c(2017, 12))
test  <- window(fcpi_log, start = c(2018, 1))

cat("Training set:", length(train), "obs |",
    "Test set:", length(test), "obs\n")
```

```{r auto-arima}
fit_auto <- auto.arima(train, ic = "aic", stepwise = FALSE, approximation = FALSE)
summary(fit_auto)
```

The `auto.arima` function selects the model that minimises the AIC using an exhaustive search over the parameter space. The selected model should be compared with the ARIMA(0,1,1)(0,1,1)$_{12}$ identified manually in 1.4. If the auto-selected model differs, this may indicate that additional AR or MA terms capture dynamics not visible from the ACF/PACF alone, or that the AIC penalisation favours a slightly different specification.

```{r residual-diagnostics, fig.height=7}
checkresiduals(fit_auto)
```

```{r ljung-box}
lb <- Box.test(residuals(fit_auto), lag = 24, type = "Ljung-Box",
               fitdf = length(fit_auto$coef))

lb_tbl <- data.frame(
  Test      = "Ljung-Box",
  Lags      = 24,
  Statistic = round(lb$statistic, 3),
  df        = lb$parameter,
  p.value   = format.pval(lb$p.value, digits = 3, eps = 0.001)
)

knitr::kable(
  lb_tbl,
  col.names = c("Test", "Lags", "X-squared", "df", "p-value"),
  caption   = "Ljung-Box test on residuals of auto.arima model.",
  align     = "lcccc"
)
```

If the Ljung-Box p-value exceeds 0.05, the residuals show no significant autocorrelation up to 24 lags and the model is adequate. The residual plot should also confirm approximate normality and constant variance. If any structure remains, adding AR or MA terms, or adjusting the seasonal component, could improve the fit.

## 2.2 Forecasting and Model Comparison

Two models are forecast over the 24 month test window: (1) the auto-selected AIC model from 2.1, and (2) the manually identified ARIMA(0,1,1)(0,1,1)$_{12}$ from 1.4.

```{r fit-manual}
fit_manual <- Arima(train, order = c(0, 1, 1), seasonal = c(0, 1, 1))
summary(fit_manual)
```

```{r forecasts}
h <- length(test)

fc_auto   <- forecast(fit_auto, h = h)
fc_manual <- forecast(fit_manual, h = h)
```

```{r forecast-plot-auto, fig.height=4.5}
autoplot(fc_auto) +
  autolayer(test, series = "Actual", size = 0.8) +
  scale_color_manual(values = c("Actual" = bene[4])) +
  scale_fill_manual(values = c(bene[2])) +
  labs(
    title = "Forecast: Auto ARIMA (AIC Selected Model)",
    x     = "Year",
    y     = "log(CPI)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )
```

```{r forecast-plot-manual, fig.height=4.5}
autoplot(fc_manual) +
  autolayer(test, series = "Actual", size = 0.8) +
  scale_color_manual(values = c("Actual" = bene[4])) +
  scale_fill_manual(values = c(bene[2])) +
  labs(
    title = "Forecast: ARIMA(0,1,1)(0,1,1)[12] (Manual Model)",
    x     = "Year",
    y     = "log(CPI)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )
```

```{r accuracy-comparison}
acc_auto   <- accuracy(fc_auto, test)
acc_manual <- accuracy(fc_manual, test)

acc_tbl <- rbind(
  data.frame(Model = "Auto ARIMA (AIC)", Set = rownames(acc_auto), acc_auto),
  data.frame(Model = "ARIMA(0,1,1)(0,1,1)[12]", Set = rownames(acc_manual), acc_manual)
)
rownames(acc_tbl) <- NULL

knitr::kable(
  acc_tbl[, c("Model", "Set", "RMSE", "MAE", "MAPE", "MASE")],
  digits  = 4,
  caption = "Forecast accuracy: Auto ARIMA vs. Manual ARIMA(0,1,1)(0,1,1)[12].",
  align   = "llcccc"
)
```

The accuracy table reports both in-sample (Training set) and out-of-sample (Test set) metrics. The model with the lower test set RMSE and MAE provides superior forecasting performance. If both models produce similar accuracy, the more parsimonious specification (fewer parameters) is preferred on grounds of interpretability and robustness. A MASE below 1 indicates that the model outperforms a naive seasonal forecast.
