---
title: "Compulsory Assignment 1"
subtitle: "Predictive Analytics (KAN-CDSCO1005U)"
author: "Alejandro Gato (160363), Alexander Myrup (161989), Pontus Neumann (185912), Linus Yu (160714)"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

\newpage

## Question 1: GDP Level -- Plot and ACF

```{r load-packages}
# install.packages("AER")
# install.packages("fpp3")
# install.packages("MetBrewer")

library(AER)
library(fpp3)
library(ggplot2)
library(MetBrewer)
```

```{r load-data}
data("USMacroG", package = "AER")
gdp <- USMacroG[, "gdp"]

# Load full Benedictus palette upfront for use throughout the assignment
bene <- met.brewer("Benedictus")

# Convert GDP to data frame for ggplot
gdp_df <- data.frame(
  date = as.Date(time(gdp)),
  gdp  = as.numeric(gdp)
)

# NBER recession periods within sample (1950-2000)
recessions <- data.frame(
  start = as.Date(c("1953-07-01", "1957-08-01", "1960-04-01", "1969-12-01",
                    "1973-11-01", "1980-01-01", "1981-07-01", "1990-07-01")),
  end   = as.Date(c("1954-05-01", "1958-04-01", "1961-02-01", "1970-11-01",
                    "1975-03-01", "1980-07-01", "1982-11-01", "1991-03-01"))
)
```

```{r gdp-plot}
ggplot(gdp_df, aes(x = date, y = gdp)) +
  geom_rect(data = recessions,
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE,
            fill = bene[4], alpha = 0.4) +
  geom_line(color = bene[1], linewidth = 0.8) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title    = "US Real GDP (1950-2000)",
    subtitle = "Shaded areas indicate NBER recession periods",
    x        = "Year",
    y        = "Billion USD"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    plot.subtitle    = element_text(color = bene[2]),
    panel.grid.minor = element_blank()
  )
```

The GDP series displays a clear upward trend throughout the entire sample period from 1950
to 2000, indicating that the series is non-stationary in its level. There is no visible
sign of seasonality, though brief slowdowns in growth are visible around the shaded
recession periods, most notably in the early 1970s and early 1980s. The variance also
appears to increase over time alongside the level of the series, which suggests that a
log transformation may be appropriate before modelling.

```{r gdp-acf}
acf_data <- acf(as.numeric(gdp), plot = FALSE)
acf_df <- data.frame(
  lag = as.numeric(acf_data$lag[-1]),
  acf = as.numeric(acf_data$acf[-1])
)
ci <- qnorm(0.975) / sqrt(length(gdp))

ggplot(acf_df, aes(x = lag, y = acf)) +
  geom_hline(yintercept = 0, color = "grey40") +
  geom_hline(yintercept = c(-ci, ci), linetype = "dashed", color = bene[3]) +
  geom_segment(aes(xend = lag, yend = 0), color = bene[1], linewidth = 0.7) +
  geom_point(color = bene[1], size = 2) +
  labs(
    title = "ACF of US Real GDP",
    x     = "Lag",
    y     = "Autocorrelation"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    panel.grid.minor = element_blank()
  )
```

The ACF confirms the non-stationarity observed in the time series plot. Autocorrelations
decay extremely slowly and remain highly significant across all displayed lags, which is
a classic symptom of a unit root or a strong deterministic trend. This persistent
autocorrelation structure further motivates transforming the data prior to any formal
modelling, as most time series models require stationarity. A log transformation is
warranted for two reasons: first, it stabilises the variance, which tends to grow
proportionally with the level of GDP; second, differencing the log series yields
approximate quarter-on-quarter percentage growth rates, which are economically
meaningful and interpretable.

\newpage

## Question 2: GDP Growth Rate -- Transformation, Plot, ACF, and Ljung-Box Test

```{r gdp-growth}
gdp_growth <- diff(log(gdp)) * 100  # Quarter-on-quarter log growth rate in %

gdp_growth_df <- data.frame(
  date   = as.Date(time(gdp_growth)),
  growth = as.numeric(gdp_growth)
)
```

```{r growth-plot}
ggplot(gdp_growth_df, aes(x = date, y = growth)) +
  geom_rect(data = recessions,
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE,
            fill = bene[4], alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(color = bene[2], linewidth = 0.7) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title    = "US Real GDP Growth Rate (QoQ)",
    subtitle = "Shaded areas indicate NBER recession periods",
    x        = "Year",
    y        = "Growth Rate (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    plot.subtitle    = element_text(color = bene[2]),
    panel.grid.minor = element_blank()
  )
```

After taking the log-difference, the series no longer exhibits a trend -- it fluctuates
around a roughly constant mean close to zero, suggesting that differencing has
successfully removed the non-stationarity present in the level series. There is no clear
seasonal pattern, though some cyclical variation is visible, with more volatile and
negative growth periods coinciding with the shaded recession episodes, most notably the
sharp contractions of the early 1970s oil shock and the double-dip recession of
1980-1982. The variance appears broadly stable across the sample, supporting the
appropriateness of the log transformation applied.

```{r growth-acf}
acf_growth <- acf(as.numeric(gdp_growth), plot = FALSE)
acf_growth_df <- data.frame(
  lag = as.numeric(acf_growth$lag[-1]),
  acf = as.numeric(acf_growth$acf[-1])
)
ci_growth <- qnorm(0.975) / sqrt(length(gdp_growth))

ggplot(acf_growth_df, aes(x = lag, y = acf)) +
  geom_hline(yintercept = 0, color = "grey40") +
  geom_hline(yintercept = c(-ci_growth, ci_growth),
             linetype = "dashed", color = bene[3]) +
  geom_segment(aes(xend = lag, yend = 0), color = bene[2], linewidth = 0.7) +
  geom_point(color = bene[2], size = 2) +
  labs(
    title = "ACF of GDP Growth Rate",
    x     = "Lag",
    y     = "Autocorrelation"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    panel.grid.minor = element_blank()
  )
```

The ACF of the growth rate decays much more rapidly than that of the level series, which
is consistent with a stationary process. A small number of lags appear statistically
significant -- notably lag 1 and potentially lag 3 -- suggesting some remaining
autocorrelation structure in the growth rate that may be relevant for model selection.
The bulk of lags, however, fall within the confidence bands, indicating that the
differenced log series behaves much closer to white noise than the original GDP series.

```{r ljung-box}
lb_test <- Box.test(gdp_growth, lag = 12, type = "Ljung-Box")

lb_tbl <- data.frame(
  Test = "Ljung-Box",
  Lags = 12,
  Statistic = round(unname(lb_test$statistic), 3),
  df        = unname(lb_test$parameter),
  p.value   = format.pval(lb_test$p.value, digits = 3, eps = 1e-3),
  check.names = FALSE
)

knitr::kable(
  lb_tbl,
  col.names = c("Test", "Lags", "X-squared", "df", "p-value"),
  caption   = "Ljung-Box test for GDP growth rate (H0: no autocorrelation up to lag 12).",
  align     = "lcccc"
)
```

The Ljung-Box test evaluates the null hypothesis that the first 12 autocorrelations are
jointly equal to zero, i.e. that the series is white noise. The test statistic is
`r round(lb_test$statistic, 3)` with a p-value of `r round(lb_test$p.value, 3)`.
Since the p-value is below the conventional 5% significance level, we reject the null
hypothesis and conclude that statistically significant autocorrelation is present in
the GDP growth rate series. This finding is consistent with the ACF plot and implies
that the growth rate is not pure white noise, which has important implications for
forecasting -- a model that exploits this autocorrelation structure should outperform a
naive no-change forecast.

\newpage

## Question 3: Olympic Running

```{r olympic-data}
data <- olympic_running
head(data)
```

```{r olympic-plot-a}
ggplot(data, aes(x = Year, y = Time, color = Sex)) +
  geom_point(size = 1.8) +
  geom_line(alpha = 0.5) +
  facet_wrap(~ Length, scales = "free_y", ncol = 3) +
  scale_color_manual(values = c("Men" = bene[1], "Women" = bene[5])) +
  labs(
    title    = "Olympic Winning Times by Event (1896-2016)",
    subtitle = "Each panel shows a track event; points are Olympic Games",
    x        = "Year",
    y        = "Winning Time (seconds)",
    color    = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    plot.subtitle    = element_text(color = bene[2]),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold")
  )
```

The plots reveal several consistent features across all events. First, there is a clear
downward trend in winning times over the course of the twentieth century, reflecting
improvements in athletic training, nutrition, equipment, and race strategy. Second, men
consistently outperform women across all distances, though women's events were introduced
later and show steeper initial declines as the sport professionalised. Third, the rate of
improvement appears to slow in more recent decades, suggesting diminishing returns as
athletes approach physiological limits. Some events also show notable outliers,
potentially reflecting unusual race conditions, doping disqualifications, or wars
causing gaps in the data (1916, 1940, 1944).

```{r olympic-regression-b}
library(dplyr)

fits <- data %>%
  group_by(Length, Sex) %>%
  filter(!is.na(Time)) %>%
  summarise(
    slope = coef(lm(Time ~ Year))[2],
    .groups = "drop"
  )

fits_tbl <- fits %>%
  mutate(`Avg. change per year (sec)` = sprintf("%.4f", slope)) %>%
  select(Event = Length, Sex, `Avg. change per year (sec)`)

knitr::kable(
  fits_tbl,
  caption = "Estimated average change in winning time per year for each Olympic event.",
  align   = "llc"
)
```

All slopes are negative, confirming that winning times have been decreasing across every
event and for both sexes. The magnitude of the slope is largest for longer distance
events in absolute terms, which is expected since longer races have more total time
available for improvement. In relative terms, however, the rate of improvement is
broadly similar across events. Women's events generally show steeper slopes, reflecting
the fact that women's athletics was less developed at the start of the observation
period and has seen larger absolute gains over time.

```{r olympic-residuals-c}
data <- data %>%
  group_by(Length, Sex) %>%
  filter(!is.na(Time)) %>%
  mutate(
    fitted   = lm(Time ~ Year)$fitted.values,
    residual = lm(Time ~ Year)$residuals
  ) %>%
  ungroup()

ggplot(data, aes(x = Year, y = residual, color = Sex)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 1.8) +
  geom_line(alpha = 0.4) +
  facet_wrap(~ Length, scales = "free_y", ncol = 3) +
  scale_color_manual(values = c("Men" = bene[1], "Women" = bene[5])) +
  labs(
    title    = "Residuals from Linear Trend by Event (1896-2016)",
    subtitle = "Residuals from OLS regression of winning time on year",
    x        = "Year",
    y        = "Residual (seconds)",
    color    = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(color = bene[1], face = "bold"),
    plot.subtitle    = element_text(color = bene[2]),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom",
    strip.text       = element_text(face = "bold")
  )
```

The residual plots reveal systematic patterns that cast doubt on the suitability of a
simple linear trend model. In most events, the residuals are not randomly scattered
around zero -- instead they tend to be positive in the early years, negative in the
middle of the sample, and positive again towards the end. This curved pattern is
consistent with a concave improvement trajectory: winning times improved rapidly in the
early decades and have slowed in recent years as athletes approach physiological limits.
A linear model cannot capture this curvature, suggesting that a nonlinear trend
(e.g. a logarithmic or quadratic specification) may provide a better fit. Additionally,
some events show clusters of large residuals around war years, reflecting the gaps in
the data.

```{r olympic-predict-d}
predictions <- data %>%
  filter(!is.na(Time)) %>%
  group_by(Length, Sex) %>%
  group_modify(~ {
    model <- lm(Time ~ Year, data = .x)
    pred  <- predict(model,
                     newdata  = data.frame(Year = 2020),
                     interval = "prediction",
                     level    = 0.95)
    data.frame(
      Predicted = round(pred[1, "fit"], 2),
      Lower     = round(pred[1, "lwr"], 2),
      Upper     = round(pred[1, "upr"], 2)
    )
  }) %>%
  ungroup()

knitr::kable(
  predictions,
  col.names = c("Event", "Sex", "Predicted (sec)", "Lower 95% PI", "Upper 95% PI"),
  caption   = "Predicted winning times for the 2020 Olympics with 95% prediction intervals.",
  align     = "llccc"
)
```

The predictions provide point estimates and 95% prediction intervals for the winning
time in each event at the 2020 Tokyo Olympics. These forecasts rest on several
important assumptions. First, the linear trend is assumed to continue into the future
at the same constant rate observed historically -- an assumption that is likely
violated, as the residual analysis above suggests the rate of improvement has been
slowing. This means the model will tend to produce optimistic (too fast) predictions,
particularly for events where the curvature in residuals is most pronounced. Second,
the model assumes that no structural breaks occur -- such as new technological
innovations, rule changes, or doping scandals -- that would shift the trend. Third,
the prediction intervals are derived under the assumption of normally distributed,
homoskedastic residuals; if this does not hold, the intervals may be too narrow or
too wide. In practice, the actual 2020 results can be compared against these
predictions to evaluate the model's out-of-sample performance.
